---
title: "Chapter 11: Clumping and Thresholding"
output:
  github_document:
    fig_width: 9
    fig_height: 5.5
    dev: png
    html_preview: false
---

```{r ch11-setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,collapse=TRUE,comment="#>",
  fig.path="figures/chapter-11-",fig.align="center",out.width="90%")
course_root <- if (file.exists("../scripts/course_helpers.R")) ".." else "."
stopifnot(file.exists(file.path(course_root,"scripts/course_helpers.R")))
source(file.path(course_root,"scripts/course_helpers.R"))
```

# Why not keep every small GWAS P value?

Nearby variants can carry overlapping information. Counting many correlated signals at full marginal weight can overemphasize one region. Clumping and thresholding (C+T) creates a simpler candidate score by selecting representatives and applying a P-value cutoff.

**Objectives:** explain clumping, distinguish it from pruning, generate candidate scores, and select a setting without consulting final test outcomes. Prerequisites: Chapters 3, 6 and 8–10. This is a complete small R demonstration, not an exact reimplementation of every PLINK clumping rule.

# 1. Two decisions

**Clumping** generally prioritizes strongly associated variants, then suppresses nearby variants sufficiently correlated with a selected index variant. **Thresholding** restricts retained variants by GWAS P value. LD pruning used for PCA usually does not prioritize GWAS association strength; it serves another purpose.

The important settings include a physical window, an LD r-squared threshold and a P-value threshold. None is universally optimal. Including weaker signals can improve a highly polygenic predictor, but it can also add noise. The best-looking setting in the test set is no longer independently evaluated. [1]

# 2. Estimate discovery associations and LD independently

```{r ch11-inputs}
demo <- make_course_data()
ss <- summary_gwas(demo)
ld_rows <- which(demo$people$role=="LD_reference")
R <- cor(demo$G[ld_rows,,drop=FALSE])
stopifnot(identical(colnames(R),ss$ID),all(is.finite(R)))
knitr::kable(head(ss),digits=4)
```

The helper fits one logistic association per variant in discovery people, adjusting for age and SBP. `beta` is the marginal log-odds coefficient conditional on these covariates; it is not the joint causal effect. `se` measures estimation uncertainty. The independent LD people describe variant correlation, not prediction accuracy.

# 3. A transparent greedy implementation

```{r ch11-clump}
clump_small <- function(ss,R,r2_cut=.1,window=250000) {
  order_p <- order(ss$p,ss$ID)
  kept <- integer(0)
  for(j in order_p) {
    nearby <- kept[ss$CHR[kept]==ss$CHR[j] & abs(ss$POS[kept]-ss$POS[j])<=window]
    redundant <- length(nearby)>0 && any(R[j,nearby]^2>r2_cut)
    if(!redundant) kept <- c(kept,j)
  }
  kept
}
kept <- clump_small(ss,R)
thresholds <- c(.001,.01,.1,1)
weights <- vapply(thresholds,function(t) {
  w <- rep(0,nrow(ss)); use <- kept[ss$p[kept]<=t]
  w[use] <- ss$beta[use]; w
},numeric(nrow(ss)))
colnames(weights) <- paste0("p_",thresholds)
knitr::kable(data.frame(P_threshold=thresholds,Variants=colSums(weights!=0)))
```

We fix the LD rule and window before examining tuning outcomes. This demonstration handles a common ordering concept but omits production details such as multiallelic handling and tool-specific secondary thresholds. Its 120 variants are not a realistic genome-wide benchmark.

# 4. Select using tuning people only

```{r ch11-tune}
it <- which(demo$people$role=="Tuning")
candidate_scores <- demo$G[it,,drop=FALSE] %*% weights
tuning_auc <- apply(candidate_scores,2,function(x) auc(demo$people$CAD5[it],x))
stopifnot(any(is.finite(tuning_auc)))
best <- which.max(tuning_auc) # Ties choose the first, more stringent threshold.
selection <- data.frame(P_threshold=thresholds,Tuning_AUC=tuning_auc,
                         Selected=seq_along(thresholds)==best)
knitr::kable(selection,digits=3)
selected_weights <- data.frame(ID=ss$ID,A1=ss$A1,weight=weights[,best])
```

An empty score ranks everyone equally and has AUC 0.5 when both outcome classes exist. It is not evidence of a usable predictor. The chosen tuning AUC is optimistic for the winner; independent testing must follow after all choices are fixed.

```{r ch11-threshold-plot, echo=FALSE, fig.cap="Figure 11.1: Tuning AUC across prespecified synthetic C+T candidates. This plot is for selection, not final performance reporting."}
plot(thresholds,tuning_auc,log="x",type="b",pch=19,col="#2874A6",
     xlab="GWAS P-value threshold",ylab="Tuning AUC")
```

# 5. Strengths, limitations and production use

C+T is relatively easy to explain and audit, and usually has lower computational demands than Bayesian genome-wide fitting. It loses information by discarding correlated variants and relies on noisy marginal coefficients. Closely linked signals and different LD patterns can change the selected set. A more sophisticated method is not guaranteed to improve performance in every dataset.

PLINK provides production clumping and score calculation as separate operations. Preserve the clump membership, retained IDs, thresholds and final weights. For a real comparison, define candidate settings beforehand, fit any clinical combination in development data, select in tuning data, and compare frozen candidates in a common independent test sample. [2]

**Published example:** the International Schizophrenia Consortium used polygenic scoring across significance thresholds to investigate shared common-variant contributions to schizophrenia and bipolar disorder. The lesson is that useful aggregate information need not be restricted to individually genome-wide significant loci; it does not imply that every weak association is causal. [3]

# Practice and handover

**Clumping versus pruning?** Clumping uses association priority; pruning usually serves a correlation-reduction goal independent of outcome association.

**Can you retune r-squared after viewing the final test result?** That makes the test part of development; a new independent evaluation is needed.

**Does this chapter find the clinically best CAD method?** No. It demonstrates selection on a small synthetic architecture.

The object `selected_weights` is an explicit learning output. To reproduce it, rerun this chapter from its helper inputs. The main evaluation chapters use the separately fixed synthetic score, so method comparisons here do not contaminate that demonstration.

# References

1. Choi SW, Mak TSH, O'Reilly PF. Tutorial: a guide to performing polygenic risk score analyses. *Nature Protocols*. 2020;15:2759–2772. https://doi.org/10.1038/s41596-020-0353-1
2. PLINK 2. Report postprocessing: clumping. https://www.cog-genomics.org/plink/2.0/postproc#clump
3. International Schizophrenia Consortium. Common polygenic variation contributes to risk of schizophrenia and bipolar disorder. *Nature*. 2009;460:748–752. https://doi.org/10.1038/nature08185
