---
title: "Chapter 8: Genotype Quality Control"
output:
  github_document:
    fig_width: 9
    fig_height: 5.5
    dev: png
    html_preview: false
---

```{r ch08-setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,collapse=TRUE,comment="#>",
  fig.path="figures/chapter-08-",fig.align="center",out.width="90%")
course_root <- if (file.exists("../scripts/course_helpers.R")) ".." else "."
stopifnot(file.exists(file.path(course_root,"scripts/course_helpers.R")))
source(file.path(course_root,"scripts/course_helpers.R"))
```

# Can poor measurements look like genetic risk?

Suppose cases were genotyped on one laboratory plate and controls on another. If the plates differ in measurement quality, a score can appear associated with disease partly because of laboratory differences. Quality control (QC) investigates whether the genetic measurements are reliable enough for their intended use.

**Objectives:** distinguish sample and variant checks, interpret missingness and allele frequency, understand why QC is iterative, and create an auditable exclusion record. Prerequisites are Chapters 2, 6 and 7. The practical uses base R and the supplied helper script; no real genotypes are required.

# 1. What QC investigates

A sample check concerns a person: unusually high missingness, unexpected heterozygosity, possible duplication, or inconsistent identifiers. A variant check concerns a locus: low call rate, unreliable imputation, unusual genotype patterns or technical differences between groups. Some findings require investigation rather than automatic deletion. [1]

**Heterozygosity** describes carrying two different alleles. An unusual value can reflect contamination, ancestry, inbreeding or a technical issue. Compare appropriate groups and markers. **Hardy–Weinberg equilibrium (HWE)** supplies expected genotype proportions under assumptions about a population. A small HWE P value is not proof of error; ancestry mixture, relatedness and disease ascertainment can matter. In case-control work, assess relevant controls and population groups rather than blindly applying one filter to everyone. [1]

| Issue | Possible consequence for PRS | Appropriate response |
|---|---|---|
| Missing genotypes | Inconsistent information across people | Inspect sample and variant missingness and scoring policy |
| Poor imputation | Dosages may be inaccurate despite being present | Inspect the imputation quality measure and source |
| Duplicates or relatives | Non-independent evaluation | Resolve duplicate samples and follow the relatedness plan |
| Batch effects | Technical variation can resemble prediction | Compare quality by laboratory batch and relevant groups |
| Population structure | Pooled summaries may conceal differences | Examine ancestry structure without equating it with nationality |

The purpose is not to make everyone genetically similar. Removing a population group merely because it differs from the largest group changes the target population and can reduce applicability.

# 2. The recurring project

Our synthetic cohort contains 120 fictional autosomal biallelic SNPs and 11,000 unrelated people: 5,000 discovery, 2,000 development, 1,000 tuning, 2,000 test and 1,000 LD-reference individuals. The last group has no supplied outcomes. Adjacent six-variant blocks share simulated haplotype information. This is a small learning model of LD, not a human reference panel.

CAD5 is an invented, fully observed five-year binary outcome. There is no censoring, competing death, ancestry structure or batch effect in the baseline simulation. The helper deliberately provides truth weights for teaching, but analytical weight estimation must not use them. A separate fixed, noisy weight vector supports the published-score-style route; it is not an actual published CAD score.

```{r ch08-recreate-data}
demo <- make_course_data()
knitr::kable(as.data.frame(table(demo$people$role)))
stopifnot(nrow(demo$G)==11000,ncol(demo$G)==120,
          identical(rownames(demo$G),demo$people$IID))
```

# 3. A worked missingness audit

We introduce failures into a copy of the development genotypes. Original simulation objects remain intact. The thresholds below are teaching decisions to make failures visible, not universal genomic QC recommendations.

```{r ch08-qc-example}
ix <- which(demo$people$role=="Development")
G <- demo$G[ix,,drop=FALSE]
G[1,1:30] <- NA                 # One person loses 25% of observations.
G[1:400,120] <- NA              # One locus loses 20% of observations.
sample_missing <- rowMeans(is.na(G))
remove_people <- sample_missing > .10
G1 <- G[!remove_people,,drop=FALSE]
variant_missing <- colMeans(is.na(G1))
remove_variants <- variant_missing > .10
G2 <- G1[,!remove_variants,drop=FALSE]
audit <- data.frame(Stage=c("Input","After person filter","After variant filter"),
                    People=c(nrow(G),nrow(G1),nrow(G2)),
                    Variants=c(ncol(G),ncol(G1),ncol(G2)))
knitr::kable(audit)
stopifnot(sum(remove_people)==1,sum(remove_variants)==1,
          nrow(G2)==1999,ncol(G2)==119)
```

We remove the person first, then recalculate locus missingness. Changing the order can change who crosses a threshold. Document the order and reassess the retained data. In a real study, the reason for missingness may be more important than the threshold itself.

```{r ch08-qc-plot, echo=FALSE, fig.cap="Figure 8.1: Artificially introduced missingness. The dashed line is a teaching threshold, not a recommended universal cutoff."}
plot(sample_missing*100,pch=16,cex=.5,col="#2874A6",
     xlab="Development participant index",ylab="Missing genotypes (%)")
abline(h=10,lty=2,col="#C0392B")
```

# 4. Frequency has a denominator

For this diploid, hard-call example, allele frequency equals the observed allele count divided by twice the number of observed genotypes. Do not count missing genotypes as zero-copy observations.

```{r ch08-frequency}
observed <- colSums(!is.na(G2))
stopifnot(all(observed>0))
freq_A <- colSums(G2,na.rm=TRUE)/(2*observed)
maf <- pmin(freq_A,1-freq_A)
summary(maf)
```

MAF means minor allele frequency: the less common of the two alleles in that sample. Minor status can differ across populations. It does not identify the scoring allele. Frequency and missingness passing does not mean dosage imputation quality has passed.

# 5. Translating to a real protocol

Start with an unchanged input trio and source metadata. Generate sample and variant summaries; inspect batch and population patterns; investigate identity and relatedness; establish documented thresholds for the actual study; apply coordinated filters; record removed IDs and reasons; recheck the remaining data. Save the retained dataset separately. Do not alter the only raw copy.

PLINK supports missingness, HWE, heterozygosity and relatedness operations, but a command list alone cannot determine appropriate thresholds. Importantly, hard-call missingness and dosage missingness are different reports. Consult the exact installed version's definitions. [2]

**Published example:** Marees and colleagues provide a GWAS QC tutorial demonstrating sample and SNP checks. It is a useful methodological example, but its choices are not universal rules for every imputed PRS dataset. [1]

# Practice and handover

**Why not replace NA with 0?** Zero is a measured allele count; NA is unknown.

**Does an HWE failure prove contamination?** No. It calls for investigation under the relevant population and sampling assumptions.

**Can we use the LD-pruned PCA markers as the entire scoring set?** Not automatically. PCA and scoring require different variant sets; retain their roles separately.

The output of this practical is the displayed `audit` and named exclusion flags. The shared cohort is recreated unchanged in later chapters. For real data, retain a sample exclusion table, variant exclusion table, QC report and versioned genotype output. Chapter 9 checks whether retained variants mean the same thing in the scoring file.

# References

1. Marees AT, de Kluiver H, Stringer S, et al. A tutorial on conducting genome-wide association studies: Quality control and statistical analysis. *International Journal of Methods in Psychiatric Research*. 2018;27:e1608. https://doi.org/10.1002/mpr.1608
2. PLINK 2. Basic statistics. https://www.cog-genomics.org/plink/2.0/basic_stats
