---
title: "Chapter 9: Genome Builds and Allele Harmonization"
output:
  github_document:
    fig_width: 9
    fig_height: 5.5
    dev: png
    html_preview: false
---

```{r ch09-setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,collapse=TRUE,comment="#>",
  fig.path="figures/chapter-09-",fig.align="center",out.width="90%")
course_root <- if (file.exists("../scripts/course_helpers.R")) ".." else "."
stopifnot(file.exists(file.path(course_root,"scripts/course_helpers.R")))
source(file.path(course_root,"scripts/course_helpers.R"))
```

# The right weight on the wrong allele

A perfectly formatted file can still produce the wrong score. If a weight belongs to allele A but the software counts allele G, the calculation answers a different question. **Harmonization** makes the identity and orientation of genetic records consistent.

**Objectives:** distinguish coordinate conversion, strand complementation and allele swapping; implement a small exact/swap audit; explain unresolved variants. Prerequisites: Chapters 2, 4, 7 and 8. The practical uses fictional diploid SNPs in base R.

# 1. A variant needs more than a label

A genome build is a version of the reference sequence used to describe positions. A coordinate on one build cannot be assumed to mean the same thing on another. An rs identifier helps identify a record but does not replace build, allele and representation checks.

An indel may have equivalent sequence representations at different neighboring positions. Normalization and reference checking help make those representations consistent. Multiallelic sites need allele-specific handling. Our exercise excludes these complexities deliberately; it is not a production harmonizer. [1]

**Effect allele** means the allele receiving the weight. **Counted allele** means the allele represented by a dosage column. **Strand complementation** replaces A with T and C with G to describe the opposite DNA strand. **Swapping** changes which of two alleles is counted. These operations are not interchangeable.

# 2. Work through three records

We assume all records share one verified fictional build, forward-strand orientation and biallelic diploid representation. This metadata is an assumption supplied by the exercise, not something the following code proves.

```{r ch09-matching}
score <- data.frame(ID=c("v1","v2","v3"),EA=c("A","C","A"),
                    OA=c("G","T","C"),weight=c(.2,-.1,.3))
target <- data.frame(ID=c("v2","v1","v3"),counted=c("T","A","G"),
                     other=c("C","G","T"),dosage=c(2,1,0))
stopifnot(!anyDuplicated(score$ID),!anyDuplicated(target$ID))
i <- match(score$ID,target$ID); stopifnot(!anyNA(i))
target <- target[i,]
exact <- score$EA==target$counted & score$OA==target$other
swap <- score$EA==target$other & score$OA==target$counted
status <- ifelse(exact,"Exact",ifelse(swap,"Swap counted allele","Unresolved"))
aligned <- rep(NA_real_,nrow(score))
aligned[exact] <- target$dosage[exact]
aligned[swap] <- 2-target$dosage[swap]
report <- data.frame(score,target_dosage=target$dosage,status,
                     effect_dosage=aligned)
knitr::kable(report)
stopifnot(identical(aligned[1:2],c(1,0)),is.na(aligned[3]))
```

For v2, two copies of T mean zero copies of C. Its original weight remains -0.1 when we transform the dosage. v3 is unresolved under our specified rules. Although a complement operation might explain its alleles in some settings, we do not invent a strand correction contrary to the stated metadata.

# 3. Why a sign flip alone changes a raw score

```text
If d counts the other allele, effect-allele dosage = 2 - d.
Correct contribution = w × (2 - d) = 2w - wd.
```

Using `-w × d` omits the constant `2w`. For a fixed set of complete diploid variants, a constant changes the raw scale without changing ranks; missingness and differing variant sets make the situation more complicated. The simple, transparent approach here is to count the documented effect allele and retain the weight.

```{r ch09-orientation-figure, echo=FALSE, fig.cap="Figure 9.1: Changing the counted allele reverses dosage direction. This identity assumes a biallelic diploid locus."}
d <- 0:2
plot(d,2-d,type="b",pch=19,lwd=2,col="#2874A6",xaxt="n",yaxt="n",
     xlab="Other-allele dosage",ylab="Effect-allele dosage",ylim=c(0,2))
axis(1,at=0:2);axis(2,at=0:2)
```

Haploid regions and some sex-chromosome analyses need different ploidy handling. Do not apply `2-d` indiscriminately across all chromosomes.

# 4. Ambiguity and missing coverage

A/T and C/G SNPs are palindromic because their allele pairs are unchanged by strand complementation. Depending on the metadata, frequency evidence and method, orientation may remain ambiguous. Similar frequencies are supporting evidence, not a universal identity test. Different populations and low-frequency alleles complicate comparisons. [2]

Do not calculate `sum(weight*aligned,na.rm=TRUE)` and present it as the complete score: that silently drops unresolved terms. Instead, report the unresolved records and apply a prespecified coverage policy. A high fraction of matched variants is reassuring only in context; missing a few influential terms can still matter.

# 5. A production handover

A real harmonization protocol starts with source metadata and confirmed builds. It normalizes appropriate variant representations, investigates duplicate keys, matches alleles, handles ambiguity according to a documented policy, and writes the final matched scoring table and exclusion report. Record exact matches, swaps, strand operations, unresolved records and absent variants separately.

Our two resolved records contribute 0.2 in total, but that is **a partial teaching sum**, not the three-variant score. The correct next action is to resolve v3 or explicitly define and validate a modified score. The course's main synthetic dataset has known A-counted genotypes and matching A/G weights; later method demonstrations use that declared orientation.

**Real application:** the PGS Catalog distributes harmonized scoring files for specified genome builds. That reduces coordinate preparation work, but the target genotypes still need compatible matching. It does not remove target QC or establish that every downloaded scoring term was used. [3]

# Practice

**Does matching the rsID prove the alleles match?** No; allele and build checks remain.

**When does 2-d apply?** To changing the counted allele at a biallelic diploid locus.

**Why keep unresolved records visible?** They explain score coverage and prevent a partial sum from being mistaken for a reproduced model.

**Can a constant shift change a percentile?** Not when the same shift is applied consistently to everyone and their reference scores. An inconsistent transformation can change interpretation.

# References

1. Danecek P, Bonfield JK, Liddle J, et al. Twelve years of SAMtools and BCFtools. *GigaScience*. 2021;10:giab008. https://doi.org/10.1093/gigascience/giab008
2. Choi SW, Mak TSH, O'Reilly PF. Tutorial: a guide to performing polygenic risk score analyses. *Nature Protocols*. 2020;15:2759–2772. https://doi.org/10.1038/s41596-020-0353-1
3. PGS Catalog. Download information. https://www.pgscatalog.org/downloads/
