---
title: "Chapter 14: Multi-Ancestry Methods and Choosing an Approach"
output:
  github_document:
    fig_width: 9
    fig_height: 5.5
    dev: png
    html_preview: false
---

```{r ch14-setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,collapse=TRUE,comment="#>",
  fig.path="figures/chapter-14-",fig.align="center",out.width="90%")
course_root <- if (file.exists("../scripts/course_helpers.R")) ".." else "."
stopifnot(file.exists(file.path(course_root,"scripts/course_helpers.R")))
source(file.path(course_root,"scripts/course_helpers.R"))
```

# Why can a large discovery study still be a poor match?

A larger GWAS can estimate associations more precisely, but differences in allele frequencies, LD and phenotype measurement can limit transfer to another population. Multi-ancestry methods attempt to combine information across discovery populations while allowing relevant differences.

**Objectives:** explain why pooling is not automatically sufficient, describe PRS-CSx, distinguish score combination from individual ancestry percentages, and choose a method using data requirements. This is an advanced chapter after Chapters 11–13.

# 1. Shared biology does not imply identical statistical inputs

A marker may tag an unobserved causal variant well in one population and poorly in another. Different allele frequencies also change available information. Environment, ascertainment and clinical care affect predictive evaluation. Broad ancestry labels simplify complex variation; they are not discrete biological categories or substitutes for local validation.

Methods differ in what they share: effect-size information, shrinkage parameters, meta-analyzed associations or final scores. A single LD panel for a mixed GWAS can be an inadequate representation. Do not average arbitrary correlation matrices and assume the resulting method is validated.

# 2. PRS-CSx and the role of tuning

PRS-CSx extends continuous-shrinkage modelling to population-specific summary statistics and LD references. One approach calculates population-specific scores and fits a combination using tuning data. Auto/meta options provide alternatives that avoid that outcome-based combination fitting, with their own assumptions. [1,2]

```text
Combined score = coefficient_1 × score_1 + coefficient_2 × score_2
```

The coefficients are prediction weights. They are not an individual's ancestry fractions. A coefficient of 0.7 does not mean 70% ancestry from a population.

# 3. A score-combination practical

This exercise creates two noisy proxies for the same synthetic score. They are **not ancestry-specific scores** and do not run PRS-CSx. The goal is to learn how combination fitting can be separated from testing.

```{r ch14-combine}
demo <- make_course_data(); d <- demo$people
set.seed(1406)
d$score1 <- d$PRS+rnorm(nrow(d),0,.3)
d$score2 <- d$PRS+rnorm(nrow(d),0,.5)
fit_rows <- d$role=="Tuning"; test_rows <- d$role=="Test"
center <- vapply(d[fit_rows,c("score1","score2")],mean,numeric(1))
spread <- vapply(d[fit_rows,c("score1","score2")],sd,numeric(1))
stopifnot(all(spread>0))
for(v in names(center)) d[[paste0(v,"_z")]] <- (d[[v]]-center[v])/spread[v]
combination <- glm(CAD5~score1_z+score2_z,data=d[fit_rows,],family=binomial())
stopifnot(combination$converged,all(is.finite(coef(combination))))
p <- predict(combination,newdata=d[test_rows,],type="response")
knitr::kable(data.frame(Coefficient=names(coef(combination)),
                        Estimate=unname(coef(combination))),digits=3)
auc(d$CAD5[test_rows],p)
```

The tuning sample fits the combination, so its performance is not final evidence. The test score here is an isolated teaching evaluation; it must not be used to select the winner of the other chapters' methods. Store the centering and scaling parameters with the coefficients when the intended use is prediction for new individuals.

Highly correlated component scores can produce unstable coefficients even when the combined prediction is reasonably stable. Negative coefficients can occur; do not interpret them as negative ancestry. Model stability and independent performance matter more than an appealing coefficient narrative.

# 4. What a real PRS-CSx run requires

Use separately documented GWAS files, their sample sizes, compatible population LD resources and target variant metadata. Check trait definition and effect scale across GWAS, and investigate overlapping discovery participants. Run each chromosome required by the model; retain posterior weights and matching reports. Decide in advance whether using fitted score combinations or a documented auto/meta route. [2]

The official repository includes an external two-population chromosome-22 test example. That is a method smoke test, not evidence for an Indian CAD cohort. Because the recurring course simulation has no population structure, it cannot validate multi-ancestry accuracy; the conceptual combination exercise above is its appropriate role.

# 5. Compare methods by the question they answer

| Approach | Main inputs | Advantage | Important limitation |
|---|---|---|---|
| Fixed published score | Weights and target genotypes | Reproducible application with fewer local choices | Relevance and coverage may be limited |
| C+T | Summary data, LD, candidate settings | Transparent, relatively economical | Discards correlated information |
| LDpred2 | Effects, SE, sample size, ordered LD | Flexible LD-aware modelling | Input mismatch and unstable fits need investigation |
| PRS-CS | Summary data and compatible block LD | Adaptive continuous shrinkage | Reference and sampling demands |
| PRS-CSx | Multiple GWAS and population references | Combines discovery information across populations | More complex compatibility and evaluation requirements |

Other families include penalized-regression methods such as lassosum and Bayesian mixture methods such as SBayesR. Their existence does not create an obligation to run every algorithm. Choose a justified comparison set, document tuning effort, and evaluate fairly. [3,4]

# 6. Published example and interpretation

Ruan and colleagues evaluated PRS-CSx in ancestrally diverse populations, providing evidence that sharing information across discovery datasets can improve prediction in evaluated settings. This is not a guarantee of equal performance across all groups or every admixed individual. [1]

A useful result in a well-represented evaluation group does not erase uncertainty in a small subgroup. Score normalization alone cannot demonstrate equal calibration or clinical utility. Chapter 19 evaluates population and setting differences explicitly.

# Practice and handover

**Are fitted combination coefficients ancestry percentages?** No; they are regression coefficients.

**Does a multi-ancestry method remove the need for target evaluation?** No.

**Why not report a competition winner from these toy chapters?** They use different didactic exercises and do not constitute a prespecified real-data benchmark.

Write a method-choice note giving your available inputs, intended population, tuning resources, computational limits and chosen comparison set. The following chapters return to the fixed synthetic score and examine what credible evaluation requires.

# References

1. Ruan Y, Lin YF, Feng YCA, et al. Improving polygenic prediction in ancestrally diverse populations. *Nature Genetics*. 2022;54:573–580. https://doi.org/10.1038/s41588-022-01054-7
2. PRS-CSx official repository. https://github.com/getian107/PRScsx
3. Mak TSH, Porsch RM, Choi SW, Zhou X, Sham PC. Polygenic scores via penalized regression on summary statistics. *Genetic Epidemiology*. 2017;41:469–480. https://doi.org/10.1002/gepi.22050
4. Lloyd-Jones LR, Zeng J, Sidorenko J, et al. Improved polygenic prediction by Bayesian multiple regression on summary statistics. *Nature Communications*. 2019;10:5086. https://doi.org/10.1038/s41467-019-12653-0
