---
title: "Chapter 10: Applying a Fixed Score"
output:
  github_document:
    fig_width: 9
    fig_height: 5.5
    dev: png
    html_preview: false
---

```{r ch10-setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,collapse=TRUE,comment="#>",
  fig.path="figures/chapter-10-",fig.align="center",out.width="90%")
course_root <- if (file.exists("../scripts/course_helpers.R")) ".." else "."
stopifnot(file.exists(file.path(course_root,"scripts/course_helpers.R")))
source(file.path(course_root,"scripts/course_helpers.R"))
```

# Reproducing a score is a distinct task

A published score supplies a model to apply. You do not need to rediscover its variants, retune its P-value threshold or estimate new weights merely to calculate it. However, you must reproduce its allele counting, weight interpretation and missing-data conventions.

**Objectives:** calculate a fixed weighted sum, standardize using a declared reference, distinguish sum from average, and prepare a reproducible scoring command. Prerequisites: Chapters 4 and 7–9.

# 1. The fixed-score route

The inputs are an identified scoring model and compatible target genotypes. Outcomes are unnecessary for arithmetic. The output is one score per person plus a coverage and processing report. A probability model needs additional coefficients and calibration information.

A **fixed score** can nevertheless be part of model development: if you estimate its coefficient alongside clinical predictors, you are fitting a new outcome model. Selecting the best published score using local outcomes also requires evaluation of the entire selection process.

# 2. Recreate and score the synthetic cohort

```{r ch10-fixed-score}
demo <- make_course_data()
G <- demo$G; w <- demo$weights
stopifnot(identical(colnames(G),w$ID),!anyDuplicated(w$ID),
          all(is.finite(G)),all(G>=0 & G<=2),all(is.finite(w$weight)))
raw_score <- as.numeric(G %*% w$weight)
stopifnot(isTRUE(all.equal(raw_score,demo$people$PRS)))
scored <- data.frame(IID=demo$people$IID,role=demo$people$role,PRS=raw_score)
knitr::kable(head(scored))
```

`%*%` multiplies a matrix by a vector. Each row is a person; each column is an A-allele count. This matches the declared A effect allele. The helper's noisy weights imitate a fixed externally supplied model but are entirely invented. Their relationship to the generating truth is a simulation convenience, not an estimate of clinical performance.

# 3. A reference transformation belongs to the model

```{r ch10-reference}
ref <- scored$PRS[scored$role=="Development"]
mu <- mean(ref); sigma <- sd(ref)
stopifnot(is.finite(sigma),sigma>0)
scored$PRS_z <- (scored$PRS-mu)/sigma
knitr::kable(data.frame(Reference="Development",Mean=mu,SD=sigma))
```

The same mean and SD are applied to test people. Standardizing the test set independently would use a different scale and would not reproduce the development transformation. This reference distribution does not by itself calibrate disease probabilities. Standardization does not guarantee a normal distribution or ancestry portability.

```{r ch10-score-distribution, echo=FALSE, fig.cap="Figure 10.1: Fixed synthetic scores standardized using development parameters. This is a score distribution, not a disease-risk distribution."}
hist(scored$PRS_z[scored$role=="Test"],breaks=30,col="#AED6F1",border="white",
     xlab="PRS in development SD units",main="Independent test participants")
```

# 4. Sum, average and missing data

PLINK 2 can output scoring sums and averages. Its average divides by the applicable allele-observation denominator, not simply by the number of variants. For complete diploid data with 120 variants that denominator is 240. Defaults can mean-impute missing genotypes; changing that policy changes the calculation. [1]

In this complete example, `raw_score/240` is a rescaled version of the sum. A positive constant rescaling preserves ranks but changes numeric units and coefficients. With variable denominators or inconsistent coverage, that simple equivalence may fail. Record the exact output column, missing-data policy and variant count. Do not assume a column labelled SCORE is the same across software versions.

# 5. Export an optional real software exercise

The companion `export_plink_demo.R` generates a small PED/MAP input and matching weights from a subset of our synthetic cohort. Run it deliberately from the course root in the Terminal; it is not run by knitting.

```bash
Rscript scripts/export_plink_demo.R
plink2 --pedmap results/plink_demo/target --make-pgen --out results/plink_demo/target
plink2 --pfile results/plink_demo/target --score results/plink_demo/weights.tsv 1 2 3 header-read cols=+scoresums --out results/plink_demo/scored
```

The script writes `expected_scores.tsv`. Compare each participant's PLINK score sum with `R_sum`, joining by IID. Small floating-point differences are normal; a tolerance of 1e-6 is adequate for this export's numerical precision. Investigate systematic disagreement before adapting the command to real data.

This exports observed hard calls, not imputed dosages. The command is a learning exercise with synthetic variants. It does not establish that a published genome-wide score was reproduced.

# 6. A real published-score route

PGS000013 is the GPS_CAD score associated with Khera et al. Its record identifies the trait, original build and model source. A full application needs the actual downloaded scoring file, compatible genome-wide target data, and a recorded matching report. [2]

Public 1000 Genomes genotypes can support a technical scoring exercise, but their public genotype files alone do not supply the prospective CAD outcomes needed for our five-year validation. Do not invent those outcomes or use the distribution of scores as proof of CAD prediction. The course README links the public resource; large downloads are optional and never triggered by knitting. [3]

# Practice and handover

**Can you score without CAD5?** Yes; calculation does not require outcomes.

**Can you exponentiate the raw sum to obtain validated disease odds?** Not automatically. That requires an appropriate model and weight interpretation.

**Why compare R and PLINK on the same synthetic data?** Agreement tests implementation choices while keeping expected arithmetic transparent.

Retain score ID, weight file, input version, score coverage, output-column definition and reference mean/SD. Chapters 11–14 introduce alternative weight-development routes. Chapters 15–20 evaluate the fixed synthetic score so the main evaluation does not secretly choose the best method on its test outcomes.

# References

1. PLINK 2. Linear scoring. https://www.cog-genomics.org/plink/2.0/score
2. Khera AV, Chaffin M, Aragam KG, et al. Genome-wide polygenic scores for common diseases identify individuals with risk equivalent to monogenic mutations. *Nature Genetics*. 2018;50:1219–1224. https://doi.org/10.1038/s41588-018-0183-z
3. 1000 Genomes Project Consortium. A global reference for human genetic variation. *Nature*. 2015;526:68–74. https://doi.org/10.1038/nature15393
