---
title: "Chapter 12: LDpred2: Modelling Correlated Genetic Effects"
output:
  github_document:
    fig_width: 9
    fig_height: 5.5
    dev: png
    html_preview: false
---

```{r ch12-setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,collapse=TRUE,comment="#>",
  fig.path="figures/chapter-12-",fig.align="center",out.width="90%")
course_root <- if (file.exists("../scripts/course_helpers.R")) ".." else "."
stopifnot(file.exists(file.path(course_root,"scripts/course_helpers.R")))
source(file.path(course_root,"scripts/course_helpers.R"))
```

# Keeping correlated information without counting it naively

Clumping makes a discrete choice about which correlated variants to keep. LDpred2 instead uses summary associations together with an LD matrix to infer adjusted weights under a statistical model of effect sizes.

**Objectives:** explain shrinkage and prior assumptions; distinguish LDpred2-inf, grid and auto; prepare ordered inputs; recognize failed candidate fits. This is an advanced method chapter after Chapters 3, 9 and 11. The input practical uses base R. An optional small software exercise requires bigsnpr, bigsparser and Matrix.

# 1. Why shrink estimates?

An estimated association combines signal and sampling noise. Shrinkage reduces the magnitude of uncertain estimates to improve prediction. It trades some bias for reduced estimation variability. A **prior** expresses a model for plausible effects before the observed summary estimates are combined with it. This is not a declaration that the biological architecture is known.

With correlated variants, individual marginal estimates partly reflect neighboring effects. An LD-aware model can distribute information across variants, but its answer depends on whether the LD matrix represents the discovery associations correctly. An accurate target ancestry label cannot rescue an incompatible discovery LD matrix. [1,2]

| Variant | Main modelling choice | Tuning outcomes |
|---|---|---|
| LDpred2-inf | Infinitesimal model with many small nonzero effects | No grid selection inherent in this variant; heritability input still needed |
| LDpred2-grid | Candidate combinations of heritability and fraction of nonzero effects; sparse options | Used to choose among candidates |
| LDpred2-auto | Learns architecture parameters using summary data | Not needed for that parameter fitting; independent evaluation still required |

Auto does not mean all QC is automatic. Multiple chains, diagnostics and sensitivity to inputs remain relevant. Candidate failures should be reported, not silently converted to zero weights.

# 2. Prepare a consistent input order

```{r ch12-prepare}
demo <- make_course_data()
ss <- summary_gwas(demo)
R <- cor(demo$G[demo$people$role=="LD_reference",,drop=FALSE])
stopifnot(identical(colnames(R),ss$ID),identical(rownames(R),ss$ID),
          all(is.finite(R)),all(ss$se>0),all(ss$n_eff>0))
df_beta <- ss[c("beta","se","n_eff")]
knitr::kable(head(df_beta),digits=4)
```

Effective sample size describes information under a chosen approximation. For our simple case-control discovery sample we use `4/(1/N_cases + 1/N_controls)`. Real meta-analysis can involve variant-specific sample sizes and more complex designs. Do not substitute total sample count for every method's requested information without checking its definition.

The matrix order and effect-allele orientation must match `df_beta`. A matrix with the right dimensions but the wrong order is still wrong. Our allele orientation is known from simulation; this code does not establish orientation for real files.

# 3. Optional small LDpred2-grid exercise

Run this R chunk deliberately after installing the named packages. It is disabled during normal knitting. The candidate heritabilities are illustrative settings, not estimates of real CAD heritability or values chosen from the hidden simulation truth.

```{r ch12-run-grid, eval=FALSE}
stopifnot(requireNamespace("bigsnpr",quietly=TRUE),
          requireNamespace("bigsparser",quietly=TRUE),
          requireNamespace("Matrix",quietly=TRUE))
# A small diagonal regularization stabilizes our estimated correlation matrix.
# Record this change; production LD preparation requires its own justification.
R_used <- .99*R + .01*diag(nrow(R))
corr <- bigsparser::as_SFBM(as(Matrix::Matrix(R_used,sparse=TRUE),"dsCMatrix"),
                           backingfile=tempfile("ld_demo_"))
params <- expand.grid(p=c(.05,.2,1),h2=c(.05,.1,.2),sparse=FALSE)
set.seed(1206)
beta_grid <- bigsnpr::snp_ldpred2_grid(corr,df_beta,params,ncores=1)
tune <- which(demo$people$role=="Tuning")
valid <- apply(beta_grid,2,function(w) all(is.finite(w)))
stopifnot(any(valid))
metric <- rep(NA_real_,ncol(beta_grid))
for(j in which(valid)) metric[j] <- auc(demo$people$CAD5[tune],
  as.numeric(demo$G[tune,,drop=FALSE] %*% beta_grid[,j]))
params$Tuning_AUC <- metric
print(params)
selected <- which.max(metric)
ldpred2_weights <- data.frame(ID=ss$ID,A1=ss$A1,weight=beta_grid[,selected])
```

Expected output is a candidate table and an ordered weight vector. Finite output is necessary but does not prove convergence, suitable architecture assumptions or useful prediction. Review diagnostics and repeat settings as justified within development; do not use the final test to decide which problematic chain to retain.

# 4. What resources and evidence are needed?

The small 120-variant exercise is an API demonstration. Genome-wide LD matrices need sparse or block storage and can consume substantial memory and disk. Never extrapolate laptop runtime from this toy exercise. The official tutorial supplies a separate educational dataset and a fuller workflow. Its tutorial phenotypes are educational, not evidence of real CAD performance. [2]

LDpred2 is useful when compatible summary data and reliable LD are available and one wants to retain information beyond clumping. It is less attractive when input scale, per-variant sample size or LD compatibility cannot be established. More modelling cannot repair wrong allele orientation.

**Published example:** Privé and colleagues benchmarked LDpred2 with simulated and real traits, including comparisons with other methods and difficult LD settings. Such benchmarking supports the method's potential, not a universal ranking across all diseases and populations. [1]

# Practice and handover

**Does auto remove the need for test outcomes?** No; it avoids outcome-based tuning of certain parameters, not independent assessment.

**Why not use the target population's LD regardless of discovery?** Summary associations encode discovery correlation patterns.

**What if no candidate is finite?** Stop and investigate inputs and model behaviour rather than manufacture a score.

Retain the reference, variant order, summary-statistic checks, parameter grid, diagnostics and chosen weights. Chapter 13 introduces a different prior family; neither chapter's toy output establishes a clinical winner.

# References

1. Privé F, Arbel J, Vilhjálmsson BJ. LDpred2: better, faster, stronger. *Bioinformatics*. 2020;36:5424–5431. https://doi.org/10.1093/bioinformatics/btaa1029
2. Privé F. Polygenic scores and inference using LDpred2. Official bigsnpr documentation. https://privefl.github.io/bigsnpr/articles/LDpred2.html
