---
title: "Chapter 1: What Is a Polygenic Risk Score?"
output:
  github_document:
    fig_width: 9
    fig_height: 5.5
    dev: png
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/chapter-01-",
  fig.align = "center",
  out.width = "90%"
)
```

# Why begin with the idea rather than the software?

Polygenic risk score analysis involves unfamiliar terms, large genetic datasets and several software tools. It is tempting to begin by running a command. That approach can produce a number, but it does not guarantee that the number is correct or meaningful.

Before calculating a score, we must understand:

- what is being added together;
- where the information came from;
- what the final score represents;
- what the score cannot tell us; and
- why the same score may not work equally well in every population.

This chapter builds that foundation.

> **Central idea:** A polygenic risk score summarizes part of a person's inherited genetic susceptibility to a trait or disease. It is a probability-related predictor, not a diagnosis, destiny or proof of causation.


# 1. Breaking down the name

The term **polygenic risk score**, abbreviated as **PRS**, contains three ideas.

| Word | Simple meaning | Scientific meaning |
|---|---|---|
| **Polygenic** | Influenced by many genetic differences | The phenotype is influenced by variants at many positions across the genome |
| **Risk** | A chance or susceptibility, not a certainty | Probability of an outcome within a defined population and time period |
| **Score** | Many pieces of information summarized as one number | A weighted sum of a person's effect-allele dosages |

The broader term **polygenic score**, or **PGS**, can be used for diseases and non-disease traits such as height. **Polygenic risk score** is commonly used when the outcome is a disease or other adverse health outcome. The literature does not always use these terms consistently, so both abbreviations will appear in scientific papers.

# 2. Why do we need a PRS?

## 2.1 Many common diseases are not controlled by one gene

Some rare disorders can be strongly influenced by a harmful variant in a single gene. Many common diseases, however, do not follow such a simple pattern.

Coronary artery disease, type 2 diabetes, breast cancer and schizophrenia are examples of **complex diseases**. Their occurrence may be influenced by:

- many genetic variants;
- age;
- lifestyle and behaviour;
- environmental exposures;
- medical history;
- social conditions; and
- chance.

For a complex disease, an individual genetic variant usually changes susceptibility only slightly. Combining information from many variants can therefore be more informative than examining one common variant at a time.

## 2.2 A classroom analogy

Imagine predicting a student's final examination performance.

Knowing how many hours the student slept last night may provide some information, but it would not tell the full story. Preparation, attendance, previous knowledge, health and examination conditions may all contribute.

A PRS works on a similar principle, but it summarizes **genetic information only**:

- each genetic variant is one small piece of information;
- some pieces receive more weight than others;
- the weighted pieces are added; and
- the total is compared with totals from other people.

The analogy has an important limitation: a PRS does not normally include lifestyle, age or clinical measurements. A complete disease-risk model may need to combine those factors with the PRS.

# 3. Where does the information come from?

Researchers first conduct a **genome-wide association study**, usually abbreviated as **GWAS**. A GWAS compares genetic variants across many people and estimates whether each variant is statistically associated with a disease or trait.

For every usable variant, the GWAS may provide an **effect estimate**, often called a *weight*. The weight describes the estimated direction and size of the variant's association with the studied outcome.

The general flow is shown below.

```{r prs-concept-diagram, echo=FALSE, fig.cap="Figure 1.1: A simplified path from a discovery GWAS to an individual's polygenic risk score."}
old_par <- par(mar = c(1, 1, 2, 1), xpd = NA)
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 7))

box <- function(xleft, ybottom, xright, ytop, label, fill) {
  rect(xleft, ybottom, xright, ytop, col = fill, border = "#2C3E50", lwd = 1.5)
  text((xleft + xright) / 2, (ybottom + ytop) / 2, label,
       cex = 1.05, font = 2)
}

box(0.3, 4.8, 2.7, 6.3, "Discovery GWAS\nMany participants", "#D6EAF8")
box(3.8, 4.8, 6.2, 6.3, "Variant weights\nEstimated effects", "#FCF3CF")
box(7.3, 4.8, 9.7, 6.3, "Target person\n0, 1 or 2 copies", "#D5F5E3")
box(3.8, 1.2, 6.2, 2.7, "Polygenic score\nOne genetic summary", "#FADBD8")

arrows(2.75, 5.55, 3.65, 5.55, length = 0.10, lwd = 2, col = "#566573")
arrows(6.25, 5.55, 7.15, 5.55, length = 0.10, lwd = 2, col = "#566573")
arrows(5.0, 4.65, 5.0, 2.9, length = 0.10, lwd = 2, col = "#566573")
arrows(8.5, 4.65, 6.35, 2.65, length = 0.10, lwd = 2, col = "#566573")

text(5, 0.45, "The score is meaningful only in a defined reference population.",
     cex = 0.95, col = "#7B241C")
par(old_par)
```

This workflow contains two different groups:

- The **discovery GWAS** estimates the variant weights.
- The **target sample** contains the people whose scores will be calculated.

Keeping these roles separate is important. If the same people are improperly used to discover, select and evaluate a score, its apparent performance may be exaggerated. We will study this problem in Chapter 7.

# 4. What is being counted?

At one genetic position, a person usually has two allele copies: one inherited from each biological parent.

For a selected **effect allele**, a person's genotype can often be coded as:

| Effect-allele dosage | Simple interpretation |
|---:|---|
| 0 | No copies of the effect allele |
| 1 | One copy of the effect allele |
| 2 | Two copies of the effect allele |

With imputed genetic data, the dosage may be a decimal such as 0.12, 0.97 or 1.84. A decimal represents uncertainty about the unobserved genotype; it does not mean that a person literally possesses a fraction of an allele.

The phrase **effect allele** is safer than automatically calling it a **risk allele**. Its weight can be:

- positive, meaning that the allele raises the score;
- negative, meaning that the allele lowers the score; or
- close to zero, meaning that its estimated contribution is small.

Furthermore, the allele used in a score is not necessarily causal. It may be correlated with another nearby variant that is closer to the biological mechanism.

# 5. The basic PRS calculation

For each variant, two values are combined:

1. the person's effect-allele dosage; and
2. the weight assigned to that variant.

The contributions are then added.

```text
Contribution of one variant = effect-allele dosage × variant weight

PRS = contribution of variant 1
    + contribution of variant 2
    + ...
    + contribution of variant M
```

Here, **M** is the number of variants included in the score.

This is a conceptual formula. The units of the score depend on how the original weights were estimated and transformed. A raw PRS should therefore not be interpreted automatically as a probability.

## 5.1 A five-variant example

Suppose a published score contains five variants. For one person, the calculation is:

```{r five-variant-table}
score_data <- data.frame(
  Variant = c("Variant 1", "Variant 2", "Variant 3", "Variant 4", "Variant 5"),
  Effect_allele = c("A", "G", "T", "C", "A"),
  Dosage = c(2, 1, 0, 2, 1),
  Weight = c(0.12, -0.06, 0.09, 0.03, 0.15)
)

score_data$Contribution <- score_data$Dosage * score_data$Weight
score_data

raw_prs <- sum(score_data$Contribution)
raw_prs
```

The same calculation written by hand is:

```text
PRS = (2 × 0.12) + (1 × -0.06) + (0 × 0.09) + (2 × 0.03) + (1 × 0.15)
    = 0.24 - 0.06 + 0.00 + 0.06 + 0.15
    = 0.39
```

The contribution of each variant can be displayed visually.

```{r variant-contributions, echo=FALSE, fig.cap="Figure 1.2: Contribution of each variant to the example person's raw PRS. A negative weight lowers the score."}
contribution_colours <- ifelse(
  score_data$Contribution >= 0,
  "#2874A6",
  "#C0392B"
)

barplot(
  score_data$Contribution,
  names.arg = score_data$Variant,
  col = contribution_colours,
  border = NA,
  ylab = "Contribution to raw PRS",
  main = "Five variants do not contribute equally",
  ylim = c(-0.10, 0.28)
)
abline(h = 0, col = "#34495E", lwd = 1.5)
```

## 5.2 What does 0.39 mean?

By itself, very little.

The value **0.39** is not automatically:

- a 39% chance of disease;
- a 39% increase in risk;
- a clinical diagnosis;
- the number of harmful variants; or
- proof that the included variants cause the disease.

To interpret it, we need information about the score's construction and its distribution in a relevant reference population.

# 6. A score becomes useful through comparison

Raw scores can differ substantially between scoring methods. One score might commonly range from -1 to 1, while another might range from -10 to 10. This does not mean that the second score is better or represents greater disease risk.

Researchers commonly transform a raw PRS into a **standardized score**:

```text
Standardized PRS = (person's raw PRS - reference mean) / reference standard deviation
```

A standardized score describes distance from the reference mean:

- 0 means equal to the reference mean;
- +1 means one standard deviation above the reference mean;
- -1 means one standard deviation below the reference mean.

A percentile gives another type of comparison. A person at the 90th percentile has a higher score than approximately 90% of people in the chosen reference group. It does **not** mean that the person has a 90% probability of disease.

```{r reference-distribution, echo=FALSE, fig.cap="Figure 1.3: A hypothetical PRS distribution. The highlighted person has a relatively high score, but the figure does not provide an absolute disease probability."}
set.seed(2026)
population_prs <- rnorm(5000, mean = 0, sd = 1)
person_prs <- 1.50
person_percentile <- mean(population_prs <= person_prs) * 100

hist(
  population_prs,
  breaks = 45,
  probability = TRUE,
  col = "#D6EAF8",
  border = "white",
  xlab = "Standardized PRS",
  main = "A PRS is interpreted relative to a reference distribution"
)
curve(dnorm(x), add = TRUE, col = "#2874A6", lwd = 2)
abline(v = person_prs, col = "#C0392B", lwd = 3, lty = 2)
text(
  person_prs,
  0.37,
  labels = paste0("Example person\nPRS = 1.50\nabout the ",
                  round(person_percentile), "th percentile"),
  pos = 4,
  col = "#922B21"
)
```

The distribution in Figure 1.3 is simulated for teaching. Real distributions depend on the score, dataset, genetic ancestry, quality-control procedures and method of standardization.

# 7. What exactly can a PRS tell us?

A properly developed and validated PRS may tell us:

- how a person's measured genetic susceptibility compares with that of a defined reference group;
- whether groups with higher scores had higher average occurrence of a disease in a validation study;
- how much predictive information the score adds to a specified model; and
- whether the score may help separate a population into broad risk groups.

The interpretation is conditional on important details:

- Which disease or trait was studied?
- Which population was used to develop the score?
- Which population was used to validate it?
- Which variants and weights were included?
- What outcome definition and follow-up period were used?
- How well was the model calibrated?
- Does the target person resemble the validation population?

# 8. What can a PRS not tell us?

## 8.1 It is not a diagnosis

A high PRS does not mean that a person currently has the disease. Diagnosis usually requires clinical history, examination, laboratory measurements, imaging or other disease-specific evidence.

## 8.2 It is not destiny

People with high PRSs may remain disease-free, while people with low PRSs may develop disease. Genetic susceptibility is only one contributor to many complex outcomes.

## 8.3 It is not usually an absolute probability

A raw or standardized PRS does not, by itself, answer:

> “What is my chance of developing this disease in the next ten years?”

Absolute-risk estimation also requires information such as baseline incidence, age, sex, competing events and sometimes clinical or environmental risk factors.

## 8.4 It does not prove causation

A PRS is mainly a prediction tool built from statistical associations. It does not prove that every included variant—or even the score itself—causes the outcome.

This distinguishes PRS from Mendelian randomization:

| Polygenic risk score | Mendelian randomization |
|---|---|
| Main purpose is prediction or risk stratification | Main purpose is causal inference |
| Combines variants to predict a trait or disease | Uses genetic variants as instruments for an exposure |
| A predictive variant need not be causal | Instruments must satisfy causal assumptions |
| Evaluated using predictive performance and calibration | Evaluated using causal estimates and sensitivity analyses |

## 8.5 It is not automatically transferable between populations

Allele frequencies, linkage disequilibrium patterns, environmental contexts and GWAS representation differ across populations. A score developed in one ancestry group may lose accuracy or become poorly calibrated in another.

This is not a minor technical issue. It is central to valid and equitable PRS analysis.

# 9. PRS, a single-gene variant and family history

These sources of information overlap, but they are not interchangeable.

| Feature | Single high-impact variant | Polygenic risk score | Family history |
|---|---|---|---|
| Genetic information | Usually one gene or variant | Many variants across the genome | Shared genes are partly represented indirectly |
| Non-genetic information | Usually not included | Usually not included | May partly reflect shared environment and behaviour |
| Typical effect | Can be large for some disorders | Each variant is usually small; combined effect may be informative | Depends on relatives, disease and family structure |
| Result | Variant classification and condition-specific interpretation | Relative genetic susceptibility or a model-based risk estimate | Observed disease pattern within a family |
| Can it replace the others? | No | No | No |

A person can therefore have:

- a high-impact pathogenic variant and a low PRS;
- no detected high-impact variant but a high PRS;
- a strong family history with an average measured PRS; or
- no family history despite elevated genetic susceptibility.

# 10. A real-life example: coronary artery disease

Coronary artery disease is influenced by many genetic and non-genetic factors. A coronary artery disease PRS may combine the effects of numerous variants to identify people with relatively higher inherited susceptibility.

In research, investigators can ask whether people in the upper tail of the PRS distribution develop coronary artery disease more often than people near the middle. The score may also be studied alongside age, blood pressure, cholesterol, smoking, diabetes and family history.

Consider two hypothetical people with equally high coronary artery disease PRSs:

| Person | Age | Smoking | Blood pressure | PRS |
|---|---:|---|---|---|
| A | 25 | No | Normal | High |
| B | 70 | Yes | High | High |

The PRS indicates high **relative genetic susceptibility** for both. It does not imply that their current or short-term absolute risks are equal. Their ages and clinical risk factors are very different.

This example shows why a PRS should complement—not replace—appropriate clinical information. Whether it improves patient outcomes must be demonstrated in the intended healthcare setting.

# 11. Why can two PRS methods give different answers?

Different methods may make different decisions about:

- which variants to include;
- how to handle correlated variants;
- how strongly to shrink uncertain effect estimates;
- which discovery GWAS to use;
- which linkage disequilibrium reference to use; and
- whether parameters are selected using a tuning dataset.

Consequently, two methods can produce different raw scores, rankings and predictive performance for the same people. A more complicated method is not automatically better. Every method must be evaluated in independent, relevant data.

Later chapters will cover:

- published fixed scores;
- clumping and thresholding;
- LDpred2;
- PRS-CS;
- penalized regression;
- SBayesR and MegaPRS;
- multi-ancestry approaches; and
- multi-trait and functionally informed approaches.

# 12. First practical protocol: inspect before calculating

Before applying any published PRS, record the following information.

## Step 1: Define the intended outcome

Write the exact disease or trait. “Heart disease” is too broad if the score was developed specifically for coronary artery disease.

## Step 2: Identify the score source

Record the publication, score identifier, scoring-file version and download date. A resource such as the PGS Catalog can provide structured information about published scores.

## Step 3: Identify the development population

Record the discovery GWAS sample size, ancestry composition, age range, sex composition and phenotype definition when available.

## Step 4: Identify the validation population

Record where the score was tested independently. Do not assume that development performance equals external performance.

## Step 5: Inspect the scoring file

At minimum, confirm that it contains:

- a variant identifier or genomic position;
- the effect allele;
- the variant weight;
- the genome build; and
- documentation of how the score was developed.

## Step 6: Ask what the output represents

Determine whether the published result is:

- a raw weighted score;
- a standardized score;
- a percentile;
- an odds ratio per standard deviation;
- a relative-risk category; or
- an absolute-risk prediction from a larger model.

## Step 7: Decide whether the score is suitable

Consider phenotype match, ancestry, genotype coverage, genome build, validation evidence and the intended use. Do not select a score simply because it contains the largest number of variants.

## A reusable inspection table

```{r inspection-template}
score_checklist <- data.frame(
  Item = c(
    "Outcome",
    "Score identifier and version",
    "Discovery GWAS",
    "Development population",
    "Independent validation population",
    "Genome build",
    "Number of variants",
    "Effect allele defined",
    "Weight scale",
    "Reported performance",
    "Intended use",
    "Important limitations"
  ),
  Information = rep("Record before analysis", 12)
)

score_checklist
```

# 13. Common misunderstandings

| Misunderstanding | Correct interpretation |
|---|---|
| “My PRS is 0.62, so my disease risk is 62%.” | A raw score is not automatically a probability. |
| “I am at the 90th percentile, so I have a 90% chance of disease.” | The percentile describes position in a reference score distribution. |
| “A high PRS means I will definitely develop the disease.” | PRS describes susceptibility, not certainty. |
| “A low PRS means I am protected.” | Low genetic susceptibility does not remove environmental, clinical or unmeasured genetic risk. |
| “All variants in my PRS cause the disease.” | Variants may be predictive because of association or correlation with nearby variants. |
| “The score worked in one cohort, so it works everywhere.” | Performance and calibration must be checked in the intended population. |
| “A statistically significant PRS must be clinically useful.” | Statistical association, prediction and clinical utility are different questions. |
| “More variants always produce a better score.” | Variant quality, weights, LD modelling, population match and validation matter more than the count alone. |

# 14. Knowledge check

## Question 1

Why are many variants combined in a PRS?

A. Every included variant causes the disease  
B. Common complex traits are often influenced by many variants with small effects  
C. A person's environment is encoded in every variant  
D. Combining variants automatically gives an absolute probability

<details>
<summary>Show answer</summary>

**Answer: B.** Many complex traits have polygenic architectures. Combining multiple small genetic contributions can capture more susceptibility than considering one common variant alone.

</details>

## Question 2

A person is at the 95th PRS percentile. What does this mean?

A. The person has a 95% chance of disease  
B. The person already has the disease  
C. The person has a higher score than approximately 95% of the selected reference group  
D. The score is 95% accurate

<details>
<summary>Show answer</summary>

**Answer: C.** A percentile describes relative position in a specified reference distribution. It is not an absolute disease probability.

</details>

## Question 3

Which statement is scientifically correct?

A. PRS proves that every included variant is causal  
B. PRS replaces clinical examination  
C. PRS is unaffected by the population used for development  
D. PRS summarizes measured genetic susceptibility and should be interpreted with population and validation information

<details>
<summary>Show answer</summary>

**Answer: D.** The validity of a PRS depends on its source, construction, validation and intended target population.

</details>

## Question 4

Suppose a person has one copy of an effect allele and the variant weight is -0.08. What is that variant's contribution?

<details>
<summary>Show answer</summary>

```text
Contribution = 1 × -0.08 = -0.08
```

The variant lowers this particular raw score by 0.08 units.

</details>

# 15. Key takeaways

1. A PRS combines information from many genetic variants into one number.
2. Each variant's contribution is its effect-allele dosage multiplied by its weight.
3. The weights usually come from GWAS results or a model trained using genetic data.
4. A raw PRS is not automatically a percentage or absolute disease probability.
5. A percentile reports position within a defined reference population.
6. A PRS predicts part of genetic susceptibility; it does not diagnose disease or prove causation.
7. Population match, ancestry, phenotype definition and independent validation are essential.
8. PRS should usually complement rather than replace clinical, environmental and family-history information.

# 16. What comes next?

Chapter 2 will introduce the genetic foundations needed for PRS: DNA, chromosomes, SNPs, alleles, genotypes, effect alleles, allele frequency and imputed dosage. No prior genetics knowledge will be assumed.

# References

1. National Human Genome Research Institute. **Polygenic Risk Scores.** National Institutes of Health. Available at: <https://www.genome.gov/Health/Genomics-and-Medicine/Polygenic-risk-scores>

2. National Human Genome Research Institute. **Polygenic Risk Score (PRS).** Talking Glossary of Genomic and Genetic Terms. Available at: <https://www.genome.gov/genetics-glossary/Polygenic-Risk-Score-PRS>

3. Choi SW, Mak TSH, O'Reilly PF. **Tutorial: a guide to performing polygenic risk score analyses.** *Nature Protocols.* 2020;15:2759–2772. <https://doi.org/10.1038/s41596-020-0353-1>

4. Lewis CM, Vassos E. **Polygenic risk scores: from research tools to clinical instruments.** *Genome Medicine.* 2020;12:44. <https://doi.org/10.1186/s13073-020-00742-5>

5. Lambert SA, Gil L, Jupp S, et al. **The Polygenic Score Catalog as an open database for reproducibility and systematic evaluation.** *Nature Genetics.* 2021;53:420–425. <https://doi.org/10.1038/s41588-021-00783-5>

6. Wand H, Lambert SA, Tamburro C, et al. **Improving reporting standards for polygenic scores in risk prediction studies.** *Nature.* 2021;591:211–219. <https://doi.org/10.1038/s41586-021-03243-6>

7. Torkamani A, Wineinger NE, Topol EJ. **The personal and clinical utility of polygenic risk scores.** *Nature Reviews Genetics.* 2018;19:581–590. <https://doi.org/10.1038/s41576-018-0018-x>

8. Martin AR, Kanai M, Kamatani Y, Okada Y, Neale BM, Daly MJ. **Clinical use of current polygenic risk scores may exacerbate health disparities.** *Nature Genetics.* 2019;51:584–591. <https://doi.org/10.1038/s41588-019-0379-x>

9. Khera AV, Chaffin M, Aragam KG, et al. **Genome-wide polygenic scores for common diseases identify individuals with risk equivalent to monogenic mutations.** *Nature Genetics.* 2018;50:1219–1224. <https://doi.org/10.1038/s41588-018-0183-z>

10. Polygenic Score Catalog. **About the PGS Catalog.** EMBL-EBI. Available at: <https://www.pgscatalog.org/about/>
